---
title: "ggplot_intro"
author: "MSCI 599"
output: html_document
---

### ggplot2: visualizing data in the tidyverse

Now we will learn how to visualise your data using `ggplot2.` R has several systems for making graphs, but `ggplot2` is one of the most elegant and most versatile. You can do more faster by learning one system and applying it to many different types of figures. Once you have gone through the basics, it can be helpful to look at a ggplot cheat sheet to remind yourself about functionality and syntax or to learn more. To download a cheat sheet, in RStudio, go to the toolbar at the top `Help -> Cheatsheets -> Data Visualization with ggplot2`. 

To use `ggplot2`', we start by loading the `tidyverse.` Note again, that you could just load the package `ggplot2` and that would also give you access to all of the `ggplot2` functions. However, by loading the `tidyverse` we can get ggplot2 and dplyr loaded at the same time (as well as some other packages that data scientists frequently like to use). Let's also load up our Antarctic penguins dataset:

```{r}
library(tidyverse)
library(palmerpenguins)
```

### Note about namespaces:

It's worth noting the messages and warnings that you get when you load in a package. In this case, we see that we are actually loading a series of packages, and all of the package versions are listed. There is also a warning about conflicts. Conflicts occur in your function namespace when you have libraries loaded with functions that have the exact same name as functions in other libraries that you loaded, or in base R. For example, there is a `filter()` function in the R `stats` package (which is automatically loaded as part of base R). There is also a `filter()` function in `dplyr`, which we used in the previous lesson. This unfortunately creates some confusion because now you may not be sure which filter() function you are actually using.

If two packages use the same function name, then the package loaded last will hide the function from earlier packages. This is called *masking*. When masking occurs and you're not sure which package you loaded last, you can just type the name of the function into the console as a command and read the output. The namespace associated with the function will indicate which package the function is coming from:

```{r}
filter
```

Let's say you actually wanted to use the `stats` package version of filter. You can specify exactly what namespace to use in front of the function like this:

```{r, eval=FALSE}
stats::filter
```

In many of my scripts, I use `dplyr`'s `select()` function to choose which columns I am interested in keeping. However, one of the other packages I frequently use has a `select()` function, so I need to specify the dplyr package before calling `select()`:

```{r}
species_islands = penguins %>% dplyr::select(species, island)
```

If you are getting unexpected behavior or output and you've triple checked your code, it may be that you aren't using the function you thought you were due to masking.

### Scatterplot

With `ggplot2`, you begin a plot with the function `ggplot()`. `ggplot()` creates a coordinate system that you can add layers to. The first argument of `ggplot()` is the dataset to use in the graph. So `ggplot(data = penguins)` creates an empty graph, but it’s not very interesting so I’m not going to show it here:

```{r, eval=FALSE}
ggplot(data=penguins)
```

You complete your graph by adding one or more layers to `ggplot()` with the addition operator `+`. The function `geom_point()` adds a layer of points to your plot, which creates a scatterplot. `ggplot2` comes with many geom functions that each add a different type of layer to a plot. We’ll learn a whole bunch of them throughout this lesson

Each geom function in `ggplot2` takes a mapping argument. The visual properties in the plot are mapped onto your dataset using the aesthetics mapping function `aes()` inside of the geom function. That's a fancy way of saying you use aes() to choose which variable in the data you want to be plotted along the x axis, and which you want to be plotted along the y axis. 

Let's use `ggplot2` to visualize the relationship between penguin body mass and penguin flipper length. By the way, I forget the names of my variables almost instantly after I look at the data. Maybe my brain is broken or something. So before I do anything, I always take a peek at my data to remind myself of the variable names and types. I usually use head() or summary(), but glimpse() is another function for peeking at your data that comes in the tidyverse:

```{r}
glimpse(penguins)
```

Now that I remember my variable names, let's build our scatterplot:

```{r}
ggplot(penguins) + 
  geom_point(aes(x=flipper_length_mm, y=body_mass_g))
```

Great! So as penguin mass goes up, so does the flipper length. Seems reasonable to me. We also got a helpful warning that 2 values were missing. That's probably because we have `NA`s in our data. We can easily check that by printing out a summary of our data:

```{r}
summary(penguins)
```

Or, if we wanted to check it more explicitly, we could count the `NA`s using `dplyr` like this:

```{r}
penguins %>%
  filter(is.na(body_mass_g)) %>% # Keep only values that are NA
  summarize(n_na = n())  # Count the NAs
```

Well the default strategy in `ggplot2` is to warn me that there are `NA`s, but plot my data anyways and just leave out the observations that are missing. That works for me. 

I wonder how the mass / flipper length relationship plays out for the three different penguin species? We can easily visualize this by making the same plot, but mapping the `color` aesthetic to species:

```{r}
ggplot(penguins) + 
  geom_point(aes(x=flipper_length_mm, y=body_mass_g, color=species))
```

Ok, so the mass vs. flipper length relationship looks pretty similar across all three species, but Gentoo penguins are bigger than Adelie and Chinstrap. That's a handy little color legend that `ggplot` made for us automatically. We'll learn how to customize that later. 

Let's make the same plot, but customize the x- and y-axis labels, and add a title to the plot. We can do this by adding more layers to our ggplot() call:

```{r}
ggplot(penguins) + 
  geom_point(aes(x=flipper_length_mm, y=body_mass_g, color=species)) + 
  xlab("Flipper length (mm)") +
  ylab("Body mass (g)") +
  ggtitle("Penguins are cute")
```

Not bad!

### Histogram

I don't really know that much about penguins. For example, how long is a penguin beak? Like a couple of centimeters? Let's plot a quick histogram and get a sense of the data:

```{r}
ggplot(penguins) + 
  geom_histogram(aes(x=flipper_length_mm))
```

Well that histogram probably has 2 peaks because the 2 smaller species are on the left and those big Gentoos are on the right. I'm just going to quickly filter the data right in the plot call to look at just the Gentoo penguin flipper length histogram:

```{r}
ggplot(penguins %>% filter(species=="Gentoo")) + 
  geom_histogram(aes(x=flipper_length_mm))
```

As I expected, that looks a bit more normal. 

I also could have colored the histogram bars by species. If I used the `color` aesthetic, that would just outline the feature with color. Instead, I'll use the `fill` aesthetic to fill the bars in with color. By specifying `alpha = 0.5` in my `geom_histogram` fucntion, I am asking the bar colors to be somewhat transparent, that way I can see if multiple colors are overlapping. Also, instead of just going with the default fill colors, I'm going to add another layer `scale_fill_manual` to specify the exact colors I want to use: 

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm)) +
  geom_histogram(aes(fill = species), alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4"))
```

Now that's pretty. See below for a bit more information on R color options. 

### bar charts

Let's build a couple of bar charts using `geom_bar` to help us get to know our data better. We can visualize the number of observations of each sex for the three different species to determine whether male and female penguins were sampled evenly:

```{r}
ggplot(data = penguins) +
  geom_bar(aes(x = sex, fill = species)) 
```

So these stacked bar charts show me that the number of males and females observed across all three species is pretty even, and there are relatively few NAs so the sex of most penguins could be identified. However, I think I would be able to interpret the figure easier if the bars were not stacked on top of each other, and if each species had it's own separate plot. The function `facet_wrap` in ggplot is a great tool for taking whatever plot you are making and splitting it into several plots according to another variable:

```{r}
ggplot(data = penguins) +
  geom_bar(aes(x = sex, fill = species)) +
  facet_wrap(~species, nrow = 1)  # Create a new plot for each species, line them all up into 1 row
```

Great. ggplot split our sex barplots into 3 separate figures, 1 for each species, and it automatically labeled the figures. This makes is really easy to see the sex distribution for each species as well as the total count across species. Bravo.

Let's make a similar set of bar plots, but this time let's visualize how many penguins of each species were observed at each of the three islands in the study. We'll use `facet_wrap()` again, since this made the information very digestible. However, since the island names are long and may not fit super nicely along the x-axis, let's try plotting the bar charts on their sides using `coord_flip()`:

```{r}
ggplot(penguins) +
  geom_bar(aes(x = island, fill = species)) +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Well, I'm glad we made that figure. Now I can easily see that Adelie were found on three islands, but Gentoo and Chinstrap were each only found on one island. That's pretty important background information if I'm going to run an analysis on these penguins. When working with new data, it's really smart to visualize it in many different ways to get to know the distributions, what's missing, and the general idiosyncracies of the dataset. Some of your exploratory graphs may be useful to put into a presentation when you are sharing your data with a group. Occassionally, these data graphs may even be important enough to publish in the papers you write.

### Saving ggplots

Since I found the species osbervation distribution across the three islands a bit surprising, I'm going to save this graph so that I can refer back to it later. I can do this by just adding one more "layer" to the ggplot to save it:

```{r}
ggplot(penguins) +
  geom_bar(aes(x = island, fill = species)) +
  facet_wrap(~species, ncol = 1) +
  coord_flip() +
  ggsave(filename = "figures/penguin_species_per_island.png", device = "png", width = 5, height = 4, units = "in", dpi = 300)
```

For this figure, I decided to save it as a `.png`, and I provided a width and height for the figure in units of inches. I also specified that I wanted the output to have a high resolution of 300 dots per inch. I didn't need to include that since `dpi=300` is the default, but now you see how to change it. Note, I also didn't need to specify "png" for the device because ggplot would have been smart enough to notice that my filename endds in ".png". I was just being very explicit here.

There are tons of options for saving your figures, and you can look through the ggsave() documentation to learn more. (Remember, do this by clicking on the `Help` tab in the lower-right hand pane of RStudio and typing ggsave into the search bar, or simply type `?ggsave` into the console).

I often save my figures as `.png` files because they have "lossless compression". That means that when I import my `.png` figures into a powerpoint and then move my powerpoint onto a different device (such as loading it onto large conference software before presenting at AGU), I don't wind up with crappy low-resolution figures. I have been burned in the past for embedding `.jpg` or `.pdf` figures. However, if I want to bring a saved figure into a different type of software for additional editing (like Adobe Illustrator), I like to save my original figures as `.pdf` because it can be easier to interact with distinct features of the graph, change fonts, etc.

### Colors

If you want a list of the named colors in R, you can use the function colors():

```{r, eval=FALSE}
colors()
```

I won't print it here because it's long. There are additional R packages and crafted color palettes available to meet whatever your aesthetic needs are. We won't dive into the nitty gritty of color options, but here is a great cheat sheet that broadly lays out a lot of the color options available in R:

https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf

### Customizing aesthetics

I used the `scale_fill_manual()` function earlier so that I could choose the exact colors that I wanted to fill in the bars of my histogram. There is a lot of functionality in this type of function (check the documentation). If you are asking yourself, "Can I change something very specific about how this figure looks?", chances are, you can. In fact, there is a whole suite of functions in ggplot that start with "scale_" followed by an aesthetic (x, y, color, fill, etc.) that you are trying to customize. Check out the `ggplot` cheat sheet to get a quick sense of the broad functionality.

### Themes

Except for customizations specific to a single aesthetic (which you use "scale_"-type fucntions for), pretty much everything that you might want to customize about your plot is done through theme(). 

